State and Parameter Estimation for Spatio-Temporal Neural Fields

### 2010-05-29 ltx/Journal of Neuroengineering/StateAndParameterEstimationNeuralFields_JoNE.tex
editted reduced model derivation and the least squares appendix

This reduced model can be manipulated into a standard state-space model in the form (see \ref{Simplifying Decomposition})

@@

	\label{Homogeneous SS Model} \mathbf{x}_{t+1} &=& q(\mathbf{x}_t)\boldsymbol{\theta} - \xi\mathbf{x}_t + \mathbf{e}_t \\
	&=& Q\left(\mathbf{x}_t\right) + \mathbf{e}_t 

@@

Q\left( \mathbf{x}_t \right) &=& q(\mathbf{x}_t)\boldsymbol{\theta} - \xi\mathbf{x}_t \label{eq:QmatrixForSigmapoints}\\
	\mathbf{e}_t &=& \label{eq:Reducednoiseterm} \boldsymbol{\Gamma}^{-1}\int_\Omega{\boldsymbol{\phi}(\mathbf{r})e_t(\mathbf{r})d\mathbf{r}}, \\
	\boldsymbol{\Gamma} &=& \int_\Omega {\boldsymbol{\phi} \left(\mathbf{r}\right)\boldsymbol{\phi} ^{\top}\left(\mathbf{r}\right)d\mathbf{r}}, \label{DefGamma} 

@@

where $\mathbf{e}_t$ is a zero-mean normal white noise term with the covariance (see~\ref{ColoredNoise}) 

@@

	\boldsymbol{\Sigma}_e=\boldsymbol{\Gamma}^{-1}\int_{\Omega}\int_{\Omega}\boldsymbol{\phi}\left(\mathbf r\right) \gamma\left(\mathbf{r}- \mathbf{r}' \right)\boldsymbol{\phi}\left(\mathbf{r}'\right)^{\top}d\mathbf{r}' d\mathbf{r}\boldsymbol{\Gamma}^{-1} 

@@

	\label{ObservationEquation} \mathbf{y}_t = \mathbf{C}\mathbf{x}_t + \boldsymbol{\varepsilon}_t,

@@

@@

\section{Reduction to Finite State-Space Model}\label{Simplifying Decomposition} 
Starting from equation~\ref{reduced continuous model} we multiply throughout by $\boldsymbol{\phi}(r)$ and integrate over the spatial domain $\Omega$ to get 
\begin{equation}\label{eq:DefGamma2}
and substituting equation~\ref{eq:DefGamma2} into equation~\ref{StartofReduction} and cross-multiplying by $\boldsymbol{\Gamma}^{-1}$ gives 
Equation~\ref{eq:ReducedForm} can be simplified by exploiting the symmetry (isotropy) of the connectivity kernel where
	\mathbf{x}_{t+1} \nonumber = \int_\Omega \boldsymbol{\Psi}(\mathbf{r}') f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t) d\mathbf{r}' \boldsymbol{\theta} + \xi\mathbf{x}_t + \boldsymbol{\Gamma}^{-1} \int_\Omega{\boldsymbol{\phi}(\mathbf{r})e_t(\mathbf{r})d\mathbf{r}}.
\dean{DEFINE q HERE!!!!!}

@@

	Consider the i.i.d. noise term $e_t\sim\mathcal{GP}(\mathbf 0,\gamma(\mathbf{s}-\mathbf{r}))$ then 
		\mathbf e_t=\boldsymbol{\Gamma}^{-1}\int_\Omega {\boldsymbol{\phi} ( \mathbf{r} )e_t( \mathbf{r} )d\mathbf{r}} \label{eq:AppendixWt} 

@@

\section{Parameter Estimation Using Least Squares}\label{LeastSquaresAppendix} 
The finite state space model  given in \ref{eq:ReducedForm} is linear in parameters and therefore Least Square (LS) method can be applied to estimate the unknown parameters. We define $n_x \times n_{\theta}$ matrix
\begin{eqnarray}\label{eq:QMatrix}
	\lefteqn {\mathbf{Q}(\mathbf{x}_t)=} \\
&&\boldsymbol{\Gamma}^{ - 1}\int_\Omega {\boldsymbol{\phi}(\mathbf{r}) \int_\Omega {f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t)\boldsymbol{\psi}^{\top} \left(\mathbf{r}-\mathbf{r}'\right)d\mathbf{r}'} d\mathbf{r}}\nonumber

@@

by substituting \ref{eq:Reducednoiseterm} and \ref{eq:QMatrix} in \ref{eq:ReducedForm} and using $\xi = 1-\zeta T_s$ at each time instant we can write
	\mathbf x_{1}&=&\mathbf x_{0}+T_s \mathbf Q(\mathbf x_0) \boldsymbol{\theta}-\zeta T_s\mathbf x_0+\mathbf e_0 \nonumber \\
	\mathbf x_{2}&=&\mathbf x_{1}+T_s \mathbf Q(\mathbf x_1) \boldsymbol{\theta}-\zeta T_s\mathbf x_1+\mathbf e_1\nonumber\\
	\mathbf x_{T}&=&\mathbf x_{T-1}+T_s \mathbf Q(\mathbf x_{T-1}) \boldsymbol{\theta}-\zeta T_s\mathbf x_{T-1}+\mathbf e_{n-1} \nonumber
% in matrix form 
% 	\left[
% 	\begin{array}{cccc}
% 		\mathbf x_{1}\\\mathbf x_{2}\\\vdots\\\mathbf x_{n}
% 	\end{array}
% 	\right] -\left[
% 	\begin{array}{cccc}
% 		\mathbf x_{0}\\\mathbf x_{1}\\\vdots\\\mathbf x_{n-1}
% 	\end{array}
% 	\right] =T_s\left[
% 	\begin{array}{cc}
% 		\mathbf Q(\mathbf x_0)&-\mathbf x_{0}\\\mathbf Q(\mathbf x_1)&-\mathbf x_{1}\\\vdots\\
% 		\mathbf Q(\mathbf x_{n-1})&-\mathbf x_{n-1}
% 	\end{array}
% 	\right] \left[
% 	\begin{array}{cc}
% 		\boldsymbol{\theta} \\
% 		\zeta
% 	\end{array}
% 	\right]+\left[
% 	\begin{array}{cccc}
% 		\mathbf e_0\\\mathbf e_1\\\vdots\\\mathbf e_{n-1}
% 	\end{array}
% 	\right] 
 in compact form we have 

@@

	\mathbf Z=\mathbf X \mathbf W+\boldsymbol \xi 

@@

		\mathbf x_{1}-\mathbf x_{0}\\
		\mathbf x_{2}-\mathbf x_{1}\\\vdots\\
		\mathbf x_{T}-\mathbf x_{T-1}

@@

	\right],\quad \mathbf X=T_s\left[

@@

		\mathbf Q(\mathbf x_0)&-\mathbf x_{0}\\
		\mathbf Q(\mathbf x_1)&-\mathbf x_{1}\\\vdots\\
		\mathbf Q(\mathbf x_{T-1})&-\mathbf x_{T-1}

@@

		\zeta

@@

		\mathbf e_0\\\mathbf e_1\\\vdots\\\mathbf e_{n-1}

@@

@@

using the LS method $ \mathbf W$ can be estimated 

@@

	\mathbf W\approx(\mathbf X^\top\mathbf X)^{-1}\mathbf X^\top\mathbf Z 

>>>>

Next we multiply equation~\ref{reduced continuous model} by $\boldsymbol{\phi}(r)$ and integrate over the spatial domain $\Omega$ to get 

@@

@@

\begin{equation}\label{eq:DefGamma}
and substituting this into equation~\ref{StartofReduction} and cross-multiplying by $\boldsymbol{\Gamma}^{-1}$ gives 

@@

This can be simplified by exploiting the symmetry (isotropy) of the connectivity kernel where

@@

	\mathbf{x}_{t+1} = \int_\Omega \boldsymbol{\Psi}(\mathbf{r}') f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t) d\mathbf{r}' \boldsymbol{\theta} + \xi\mathbf{x}_t + \boldsymbol{\Gamma}^{-1} \int_\Omega{\boldsymbol{\phi}(\mathbf{r})e_t(\mathbf{r})d\mathbf{r}}.
Now we define the state disturbance as
\begin{equation}\label{eq:AppendixWt} 
	\mathbf{e}_t=\boldsymbol{\Gamma}^{-1}\int_\Omega {\boldsymbol{\phi} ( \mathbf{r} )e_t( \mathbf{r} )d\mathbf{r}} 
which is a zero mean normally distributed white noise process with covariance (see~\ref{ColoredNoise})
	\boldsymbol\Sigma_e =\mathbf{\Gamma}^{-1}\int_{\Omega}\int_{\Omega}\boldsymbol{\phi}\left(\mathbf r\right) \gamma\left(\mathbf r- \mathbf r' \right)\boldsymbol{\phi}\left(\mathbf r'\right)^{\top}d\mathbf r' d\mathbf r\mathbf{\Gamma}^{- \top}. 

@@

\begin{equation}\label{ObservationEquation} 
	\mathbf{y}_t = \mathbf{C}\mathbf{x}_t + \boldsymbol{\varepsilon}_t,

@@

Now we have the final form of the state-space model where
	\mathbf{x}_{t+1} = \int_\Omega \boldsymbol{\Psi}(\mathbf{r}') f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t) d\mathbf{r}' \boldsymbol{\theta} + \xi\mathbf{x}_t +\mathbf{e}_t.
	\mathbf{y}_t = \mathbf{C}\mathbf{x}_t + \boldsymbol{\varepsilon}_t,

This reduced model can be manipulated into a standard state-space model in the form (see \ref{Simplifying Decomposition})
\begin{eqnarray}
	\label{Homogeneous SS Model} \mathbf{x}_{t+1} &=& q(\mathbf{x}_t)\boldsymbol{\theta} - \xi\mathbf{x}_t + \mathbf{e}_t \\
	&=& Q\left(\mathbf{x}_t\right) + \mathbf{e}_t 
\end{eqnarray}
where
\begin{eqnarray}
Q\left( \mathbf{x}_t \right) &=& q(\mathbf{x}_t)\boldsymbol{\theta} - \xi\mathbf{x}_t \label{eq:QmatrixForSigmapoints}\\
	\mathbf{e}_t &=& \label{eq:Reducednoiseterm} \boldsymbol{\Gamma}^{-1}\int_\Omega{\boldsymbol{\phi}(\mathbf{r})e_t(\mathbf{r})d\mathbf{r}}, \\
	\boldsymbol{\Gamma} &=& \int_\Omega {\boldsymbol{\phi} \left(\mathbf{r}\right)\boldsymbol{\phi} ^{\top}\left(\mathbf{r}\right)d\mathbf{r}}, \label{DefGamma} 
\end{eqnarray}

@@

% \section{Reduction to Finite State-Space Model}\label{Simplifying Decomposition} 
% Starting from equation~\ref{reduced continuous model} we multiply throughout by $\boldsymbol{\phi}(r)$ and integrate over the spatial domain $\Omega$ to get 
% \begin{eqnarray}
% 	\label{StartofReduction}\lefteqn{ \int_\Omega {\boldsymbol{\phi} \left(\mathbf{r}\right)\boldsymbol{\phi}^{\top}\left(\mathbf{r}\right) d\mathbf{r}} \mathbf{x}_{t+1}=} \nonumber\\
%  &&T_s \int_\Omega {\boldsymbol{\phi} (\mathbf{r}) \boldsymbol{\theta}^{\top} \int_\Omega {\boldsymbol{\psi} \left(\mathbf{r}-\mathbf{r}'\right) f\left(\boldsymbol{\phi}^{\top}\left(\mathbf{r}'\right) \mathbf{x}_t \right)d\mathbf{r}'}d\mathbf{r}} \nonumber\\
% &+& \xi\int_\Omega {\boldsymbol{\phi}(\mathbf{r})\boldsymbol{\phi}^{\top}(\mathbf{r})d\mathbf{r}} \mathbf{x}_t+
% \int_\Omega{\boldsymbol{\phi} \left(\mathbf{r}\right) e_t\left(\mathbf{r}\right)d\mathbf{r}}. 
% \end{eqnarray}
% Now defining
% \begin{equation}\label{eq:DefGamma2}
% 	\boldsymbol{\Gamma} = \int_\Omega {\boldsymbol{\phi} \left(\mathbf{r}\right)\boldsymbol{\phi} ^{\top}\left(\mathbf{r}\right)d\mathbf{r}}, 
% and substituting equation~\ref{eq:DefGamma2} into equation~\ref{StartofReduction} and cross-multiplying by $\boldsymbol{\Gamma}^{-1}$ gives 
% \begin{eqnarray}\label{eq:ReducedForm}
% 	 \lefteqn{\mathbf{x}_{t+1} \nonumber = T_s\boldsymbol{\Gamma}^{ - 1}\int_\Omega {\boldsymbol{\phi}(\mathbf{r}) \int_\Omega {f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t) \boldsymbol{\psi}^{\top} (\mathbf{r}-\mathbf{r}')d\mathbf{r}'} d\mathbf{r}} \boldsymbol{\theta}} \nonumber\\
% &+&\xi\mathbf{x}_t + \boldsymbol{\Gamma}^{-1} \int_\Omega{\boldsymbol{\phi}(\mathbf{r})e_t(\mathbf{r})d\mathbf{r}}.
% \end{eqnarray}
% Equation~\ref{eq:ReducedForm} can be simplified by exploiting the symmetry (isotropy) of the connectivity kernel where
% 	\boldsymbol{\psi} (\mathbf{r}-\mathbf{r}') = \boldsymbol{\psi} (\mathbf{r}'-\mathbf{r}).
% To make the simplification we first define
% \begin{equation}\label{eq:DefPsi}
% 	\boldsymbol{\Psi}(\mathbf{r}') = T_s\boldsymbol{\Gamma}^{-1}\int_\Omega {\boldsymbol{\phi}(\mathbf{r})\boldsymbol{\psi} (\mathbf{r}'-\mathbf{r})d\mathbf{r}},
% which is constant and can be defined analytically. Now substituting~\ref{eq:DefPsi} into~\ref{eq:ReducedForm} gives
% 	\mathbf{x}_{t+1} \nonumber = \int_\Omega \boldsymbol{\Psi}(\mathbf{r}') f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t) d\mathbf{r}' \boldsymbol{\theta} + \xi\mathbf{x}_t + \boldsymbol{\Gamma}^{-1} \int_\Omega{\boldsymbol{\phi}(\mathbf{r})e_t(\mathbf{r})d\mathbf{r}}.
% \dean{DEFINE q HERE!!!!!}

@@

	Consider the i.i.d. noise term $e_t(\mathbf{r})\sim\mathcal{GP}(\mathbf 0,\gamma(\mathbf{r}-\mathbf{r'}))$ then 
	\begin{equation}\label{eq:AppendixWt} 
		\mathbf e_t=\boldsymbol{\Gamma}^{-1}\int_\Omega {\boldsymbol{\phi} ( \mathbf{r} )e_t( \mathbf{r} )d\mathbf{r}} 

@@

\section{Parameter Estimation Using Least Squares}\label{LeastSquaresAppendix} 
The finite state space model  given in \ref{eq:ReducedForm} is linear in parameters and therefore Least Square (LS) method can be applied to estimate the unknown parameters. We define $L \times n_{\theta}$ matrix
\begin{eqnarray}\label{eq:QMatrix}
	\lefteqn {\mathbf{Q}(\mathbf{x}_t)=} \\
&&\boldsymbol{\Gamma}^{ - 1}\int_\Omega {\boldsymbol{\phi}(\mathbf{r}) \int_\Omega {f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t)\boldsymbol{\psi}^{\top} \left(\mathbf{r}-\mathbf{r}'\right)d\mathbf{r}'} d\mathbf{r}}\nonumber
\end{eqnarray}
by substituting \ref{eq:Reducednoiseterm} and \ref{eq:QMatrix} in \ref{eq:ReducedForm} and using $\xi = 1-\zeta T_s$ at each time instant we can write
\begin{eqnarray}
	\mathbf x_{1}&=&\mathbf x_{0}+T_s \mathbf Q(\mathbf x_0) \boldsymbol{\theta}-\zeta T_s\mathbf x_0+\mathbf e_0 \nonumber \\
	\mathbf x_{2}&=&\mathbf x_{1}+T_s \mathbf Q(\mathbf x_1) \boldsymbol{\theta}-\zeta T_s\mathbf x_1+\mathbf e_1\nonumber\\
	&\vdots& \\
	\mathbf x_{T}&=&\mathbf x_{T-1}+T_s \mathbf Q(\mathbf x_{T-1}) \boldsymbol{\theta}-\zeta T_s\mathbf x_{T-1}+\mathbf e_{T-1} \nonumber
\end{eqnarray}
% in matrix form 
% \begin{equation}
% 	\left[
% 	\begin{array}{cccc}
% 		\mathbf x_{1}\\\mathbf x_{2}\\\vdots\\\mathbf x_{n}
% 	\end{array}
% 	\right] -\left[
% 	\begin{array}{cccc}
% 		\mathbf x_{0}\\\mathbf x_{1}\\\vdots\\\mathbf x_{n-1}
% 	\end{array}
% 	\right] =T_s\left[
% 	\begin{array}{cc}
% 		\mathbf Q(\mathbf x_0)&-\mathbf x_{0}\\\mathbf Q(\mathbf x_1)&-\mathbf x_{1}\\\vdots\\
% 		\mathbf Q(\mathbf x_{n-1})&-\mathbf x_{n-1}
% 	\end{array}
% 	\right] \left[
% 	\begin{array}{cc}
% 		\boldsymbol{\theta} \\
% 		\zeta
% 	\end{array}
% 	\right]+\left[
% 	\begin{array}{cccc}
% 		\mathbf e_0\\\mathbf e_1\\\vdots\\\mathbf e_{n-1}
% 	\end{array}
% 	\right] 
% \end{equation}
 in compact form we have 
\begin{equation}
	\mathbf Z=\mathbf X \mathbf W+\boldsymbol \xi 
\end{equation}
where 
\begin{small}
\begin{equation*}
	\mathbf Z=\left[
	\begin{array}{cccc}
		\mathbf x_{1}-\mathbf x_{0}\\
		\mathbf x_{2}-\mathbf x_{1}\\\vdots\\
		\mathbf x_{T}-\mathbf x_{T-1}
	\end{array}
	\right],\quad \mathbf X=T_s\left[
	\begin{array}{cccc}
		\mathbf Q(\mathbf x_0)&-\mathbf x_{0}\\
		\mathbf Q(\mathbf x_1)&-\mathbf x_{1}\\\vdots\\
		\mathbf Q(\mathbf x_{T-1})&-\mathbf x_{T-1}
	\end{array}
	\right] 
\end{equation*}
\end{small}
and
\begin{small}
\begin{equation*}
\quad \mathbf W=\left[
	\begin{array}{cc}
		\boldsymbol{\theta} \\
		\zeta
	\end{array}
	\right],\quad \boldsymbol \xi=\left[
	\begin{array}{cccc}
		\mathbf e_0\\\mathbf e_1\\\vdots\\\mathbf e_{n-1}
	\end{array}
	\right] 
\end{equation*}
\end{small}
using the LS method $ \mathbf W$ can be estimated 
\begin{equation}
	\mathbf W\approx(\mathbf X^\top\mathbf X)^{-1}\mathbf X^\top\mathbf Z 
\end{equation}

\section{Parameter Estimation Using Least Squares - Dean}\label{App:LeastSquaresAppendix} 
The finite state space model given in \ref{eq:ReducedForm} is linear with respect to parameters. Therefore a least square (LS) method can be applied to estimate the unknown parameters. We define $L \times n_{\theta}$ matrix
	\mathbf{q}(\mathbf{x}_t) = \int_\Omega \boldsymbol{\Psi}(\mathbf{r}') f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t) d\mathbf{r}' \boldsymbol{\theta}
Given a state estimate sequence from the unscented Rauch Tung Striebel smoother (URTSS), we can write
	\mathbf x_{1} &=& \mathbf{q}(\mathbf x_0) \boldsymbol{\theta}+\xi\mathbf x_0+\mathbf e_0 \nonumber \\
	\mathbf x_{2} &=& \mathbf{q}(\mathbf x_1) \boldsymbol{\theta}+\xi\mathbf x_1+\mathbf e_1 \nonumber \\
	\mathbf x_{T}&=&\mathbf{q}(\mathbf x_{T-1}) \boldsymbol{\theta}+\xi\mathbf x_{T-1}+\mathbf e_{T-1}. \nonumber \\

@@

This can be written in the compact form

@@

	\mathbf Z=\mathbf X \mathbf W+\mathbf{e} 

@@

		\mathbf x_{1}\\
		\mathbf x_{2}\\
		\vdots\\
		\mathbf x_{T}

@@

	\right],\quad \mathbf X=\left[

@@

		\mathbf q(\mathbf x_0)& \mathbf x_{0}\\
		\mathbf q(\mathbf x_1)& \mathbf x_{1}\\
		\vdots & \vdots\\
		\mathbf q(\mathbf x_{T-1})& \mathbf x_{T-1}

@@

		\xi

@@

		\mathbf e_0\\
		\mathbf e_1\\
		\vdots\\
		\mathbf e_{T-1}

@@

	\right].

@@

Following this, the parameters, $ \mathbf W$, can be estimated by

@@

	\mathbf W\approx(\mathbf X^\top\mathbf X)^{-1}\mathbf X^\top\mathbf Z. 
